Teaching modeling in introductory statistics: A comparison of formula and tidyverse syntaxes

### 2024-01-24 PaperDraft.Rmd
spellcheck
(hopefully) final revisions

abstract: >
  This paper reports on a head-to-head comparison run in a pair of introductory statistics labs, one conducted fully in the formula syntax, the other in tidyverse. Analysis of pre- and post-survey data show no differences between the two labs, so students appeared to have a positive experience regardless of section. Analysis of incidental data from YouTube and RStudio Cloud show interesting distinctions. The formula section appeared to watch a larger proportion of pre-lab YouTube videos, but spend less time computing on RStudio Cloud. Conversely, the tidyverse section watched a smaller proportion of the videos and spent more time computing. Analysis of lab materials showed that tidyverse labs tended to be slightly longer in terms of lines in the provided RMarkdown materials and minutes of the associated YouTube videos. Both labs relied on a quite small vocabulary of consistent functions, which can provide a starting point for instructors interested in teaching introductory statistics in R. The tidyverse labs exposed students to slightly more distinct R functions, but reused functions more frequently. This work provides additional evidence for instructors looking to choose between syntaxes for introductory statistics teaching. 
bibliography: references.bib
output:
  bookdown::pdf_book:
    base_format: rticles::asa_article
    keep_tex: yes
    includes:
      in_header: header.tex
    citation_package: natbib
header-includes:
-  \usepackage{amsmath}
editor_options: 
  chunk_output_type: console
---

@@

A comparison of using the three syntaxes for univariate statistics and displays can be seen in \ref{r-syntax}. This example code, like the rest in this paper, uses the `palmerpenguins` data [@horstetal2020]. All three pieces of code accomplish the same tasks, and all three use the R language. But, the syntax varies considerably.

@@

In Figure \ref{fig:pre-post}, you can see a visualization of these Likert-scale questions.The visualization shows the pre- and post-survey responses to the survey questions, broken down by section. There is no obvious overall trend. However, considerable improvement was seen in in the Programming Confident category, which suggests that students in both sections became more confident in their "ability to make use of programming software to work with data."

@@

In Figure \ref{fig:prepost}, many of the boxplots are centered 0 (meaning the median response did not change over the course of the semester), so there was no overall difference in medians for those questions.

@@

Regardless of the syntax they choose, we recommend instructors attempt to reduce the number of functions they expose students to over the course of a semester, particularly in an introductory class, and repeat them as much as possible. This will help reduce cognitive load. If you are optimizing for the minimum number of functions, the formula syntax may be the best solution. If you would prefer to maximimize repetition, the `tidyverse` syntax appears better.

>>>>

abstract: >
  This paper reports on a head-to-head comparison run in a pair of introductory statistics labs, one conducted fully in the formula syntax, the other in tidyverse. Analysis of pre- and post-survey data show minimal differences between the two labs, with students reporting a positive experience regardless of section. Analysis of data from YouTube and RStudio Cloud show interesting distinctions. The formula section appeared to watch a larger proportion of pre-lab YouTube videos, but spend less time computing on RStudio Cloud. Conversely, the tidyverse section watched a smaller proportion of the videos and spent more time computing. Analysis of lab materials showed that tidyverse labs tended to be slightly longer in terms of lines in the provided RMarkdown materials and minutes of the associated YouTube videos. Both labs relied on a relatively small vocabulary of consistent functions, which can provide a starting point for instructors interested in teaching introductory statistics in R. The tidyverse labs exposed students to more distinct R functions, but reused functions more frequently. This work provides additional evidence for instructors looking to choose between syntaxes for introductory statistics teaching. 
bibliography: references.bib
biblio-style: apalike
output:
  bookdown::pdf_book:
    base_format: rticles::asa_article
    keep_tex: yes
    includes:
      in_header: header.tex
    citation_package: natbib
header-includes:
-  \usepackage{amsmath}
editor_options: 
  chunk_output_type: console
---

@@

A comparison of using the three syntaxes for univariate statistics and displays can be seen in \ref{r-syntax}. This example code, like the rest in this paper, uses the `palmerpenguins` data [@horstetal2020]. All three pieces of code accomplish the same tasks, and all three use the R language. However, the syntax varies considerably.

@@

Figure \ref{fig:pre-post} displays a visualization of these Likert-scale questions.The visualization shows the pre- and post-survey responses to the survey questions, broken down by section. There is no obvious overall trend. However, considerable improvement was seen in in the Programming Confident category, which suggests that students in both sections became more confident in their "ability to make use of programming software to work with data."

@@

In Figure \ref{fig:prepost}, many of the boxplots are centered at zero (meaning the median response did not change over the course of the semester), so there was no overall difference in medians for those questions.

@@

Regardless of the syntax they choose, we recommend instructors attempt to reduce the number of functions they expose students to over the course of a semester, particularly in an introductory class, and repeat them as much as possible. This will help reduce cognitive load. If you are optimizing for the minimum number of functions, the formula syntax may be the best solution. If you would prefer to maximize repetition, the `tidyverse` syntax appears better.
